---
title: "MB5370_Mod4_WS3"
author: "Simi Komenda"
date: "2024-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Workshop 3 - Data Wrangling 
tidy data - take tabular data and prepare it for subsequent use

## Install and load tidyverse packages 
```{r install and load tidyverse}
library(tidyverse)
library(dplyr)
```

## Data Wrangling
tibble is a data frame in tidyr. 
%>% is a pipe to better understand what the code is doing. it takes the data left of the pipe and applies the function right of the pipe (|> acheives the same thing)

## Import data
link to QFISH data: https://qfish.fisheries.qld.gov.au/query/number-caught-by-area-calendar-year-and-species-group/table# . 
Fix raw data in excel before importing into R (original data is a mess and R doesn't like it)
```{r QFISH}
#save QFISH data as CSV and add CSV to data folder and import into R
QFISH <- read_csv("../data/QFISH.csv")
QFISH <- mutate_all(QFISH, ~replace(., is.na(.),0)) # replace NA with 0
```

## Tidy QFISH data 
tidy the data year by year (runs from 2001-2024 but I only did five years due to time constraints)

2001
```{r tidy 2001}
yr2001 <- QFISH |>
  dplyr::select(Area, starts_with("2001")) #select the year 2001 first 

colnames(yr2001) <- unlist(yr2001[1,]) #removes 1st row  - year

yr2001 <- 
  yr2001 |>
  slice(-1) |> # removes repeated catch row 
    rename(Area = Catch) |> #replace the header catch with area
  mutate(Year = 2001) #adds the row year and fills it with 2001
```

2002
```{r tidy 2002}
yr2002 <- QFISH |>
  dplyr::select(Area, starts_with("2002")) #select the year 2002  

colnames(yr2002) <- unlist(yr2002[1,]) #removes the 1st row - year
yr2002 <- 
  yr2002 |>
  slice(-1) |> # removes repeated catch row
    rename(Area = Catch) |> #renames the cell to area
  mutate(Year = 2002) #adds the row year and fills it with 2002

```

2003
```{r tidy 2003}
yr2003 <- QFISH |>
  dplyr::select(Area, starts_with("2003")) #select the year 2002  

colnames(yr2003) <- unlist(yr2003[1,]) #removes the 1st row - year
yr2003 <- 
  yr2003 |>
  slice(-1) |> # removes repeated catch row
    rename(Area = Catch) |> #renames the cell area
  mutate(Year = 2003) #creates the column year and fills it with 2003

```

2004
```{r tidy 2004}
yr2004 <- QFISH |>
  dplyr::select(Area, starts_with("2004")) #select the year 2004  

colnames(yr2004) <- unlist(yr2004[1,]) #removes the 1st row - year
yr2004 <- 
  yr2004 |>
  slice(-1) |> # removes repeated catch row
    rename(Area = Catch) |> #renames the cell area
  mutate(Year = 2004) #creates the column year and fills it with 2004
```

2005
```{r tidy 2005}
yr2005 <- QFISH |>
  dplyr::select(Area, starts_with("2005")) #select the year 2005  

colnames(yr2005) <- unlist(yr2005[1,]) #removes the 1st row - year
yr2005 <- 
  yr2005 |>
  slice(-1) |> # removes repeated catch row
    rename(Area = Catch) |> #renames the cell area
  mutate(Year = 2005) #creates the column year and fills it with 2005

```

## Join the tables together 
bind the 5 tables on top of each other 

```{r join tables}
QFISH_1 <- 
  yr2001 |> 
  bind_rows(yr2002,yr2003,yr2004,yr2005) |> # joins rows with the same name 
  
  #needed to make the y axis continuous - mutate variables from chr to numeric 
  mutate(Area = as.factor(Area)) |> # convert the area variable to factor (no longer chr)
  mutate_if(is.character, as.numeric) #converts the other variables to numeric 
```

## Plot the Data
create a time series for the QFISH data 
```{r QFISH plot}
ggplot(data = QFISH_1) +
  geom_point(mapping = aes(x = Year, y = Total, color = Area)) + #plots points
  geom_line(mapping = aes(x = Year, y = Total, color = Area)) + #plots line
  scale_y_continuous(breaks = seq(0, 180, by = 20)) + #sets scale of y axis 
  theme_classic() + #theme with no background or grid lines 
  labs(x = "Calendar Year", y = "Total Catch", #label axis
       caption = "Figure 3. Total catch (mammals, tutrles, sharks, and others) per calendar year.") #add caption
```

